---
title: "Camila Rodriguez Lab9, ENVIRON 710"
output:
  html_document: default
  pdf_document: default
---

## Setup

```{r message=FALSE, warning=FALSE}
#install.packages("DHARMa")
#install.packages("gtsummary")
library(gtsummary)

library(DHARMa)
library(here)
library(tidyverse)

lizards<- read.csv('jrn_lizard.csv')

lizards<- lizards %>% filter(spp== "UTST")

#unique(lizards$zone)
```

### Research Question: Do snout to vent length, sex, and vegetation zone at time of capture significantly predict if a lizard tail is recorded as whole?

## Data Visualization

```{r}

# Plot 1: Bar Plot of Lizards by Zone
p1 <- ggplot(lizards, aes(x = zone)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(title = "Distribution of Lizards by zones", x = "Zone", y = "Frequency")

# Plot 2: Sex Distribution
#filter out N/A sexes 

lizards <- lizards %>%
  filter(sex %in% c("F", "M", "J") & !is.na(sex))


p2 <- ggplot(lizards, aes(x = sex)) +
  geom_bar(binwidth = 1, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Lizards by Sex ", x = "Sex", y = "Frequency")

# Plot 3: Snout to vent length
p3 <- ggplot(lizards, aes(x = SV_length)) +
  geom_histogram(fill = "coral", color = "black") +
  labs(title = "Lizard distribution of snout to vent lengths", x = "Snout to Vent Length ", y = "Number of Observations") 
#left skewed

p4<- ggplot(lizards, aes(x=zone, y=SV_length))+ geom_boxplot()

p5<- ggplot(lizards, aes(x=sex, y=SV_length))+ geom_boxplot()



# View plots and analyze possible correlations
p1
p2
p3
p4
p5

```

## Fit Regresison

```{r}

#convert necessary data into factors for GLM

lizards$sex <- factor(lizards$sex, levels= c("J", "F", "M"))
#unique(lizards$sex)

lizards$zone <- factor(lizards$zone, levels= c("C", "G", "M", "T"))
#unique(lizards$zone)

#filter out NAs from tails, make into factor 
lizards <- lizards %>%
  filter(tail %in% c("B", "W")) %>%
  mutate(tail = factor(tail, levels = c("B", "W")))

tail_model<- glm(tail~ sex + zone+ SV_length, data= lizards, family= "binomial")
```

## Evaluate Model Diagnostics

The results of a logistic regression suggest that the log-odds of having a tail are significantly lower in zone G (β = -0.57, p = 0.001) and zone T (β = -0.40, p = 0.025) relative to the reference zone (C). There was also a significant negative relationship between SV length and the log-odds of having a tail (β = -0.045, p \< 0.001). Although zone M had a negative coefficient (β = -0.30), this effect was marginally non-significant (p = 0.063). Neither sex (female: β = -0.20, p = 0.32; male: β = -0.14, p = 0.53) was significantly associated with the log-odds of having a tail.

```{r}
summary(tail_model)
simulateResiduals(tail_model) %>% plot()
```

##Probability Graphs for Zones and Sexes, Snout to Vent Length

```{r}
#graph for sex

# Simulate SV length. Range is 0-87 so we choose 90 to round, and 3 (for three sexes)
SV_length_vector <- rep(seq(from = 0, to = 90), 3)

# Simulate sex data repeating "F", "J", "M"
sex_vector <- rep(c("F", "J", "M"), each = length(seq(from = 0, to = 90)))

# Keep zone "C" constant
zone_vector <- rep("C", length(SV_length_vector))

# Combine into data frame
data_pred_sex <- data.frame(SV_length_vector, zone_vector, sex_vector)
colnames(data_pred_sex) <- c("SV_length", "zone", "sex")

# Predict using model
prediction_sex<- predict(tail_model,
                      newdata = data_pred_sex,
                      type = "response",
                      se.fit = TRUE)

# Create final data frame with predictions
data_fig <- data.frame(data_pred_sex,
                       prediction_sex$fit,
                       prediction_sex$se.fit)

# Rename columns
colnames(data_fig) <- c("SV_length", "zone", "sex", "probability", "se")

# Graph probab. of a lizard tail being whole (W) across lizard lengths,  colored by sex
ggplot(data_fig, aes(x = SV_length, y = probability)) +
  geom_line(aes(color = sex)) +
  geom_ribbon(aes(ymin = probability - se,
                  ymax = probability + se,
                  fill = sex), alpha = 0.4) +
  labs(x = "SV Length",
       y = "Probability of Having Whole Tail",
       color = "Sex",
       fill = "Sex") +
  theme_bw() +
  ggtitle("Probability of Whole Tail with Change in Lizard Length by Sex")

## ------Repeat for zone------

# Simulate SV length
SV_length_vector_2 <- rep(seq(from = 0, to = 90), 4)

# Keep sex constant at "J"
sex_vector_2 <- rep("J", length(SV_length_vector_2))

# Simulate zone data repeating "F", "J", "M"
zone_vector_2 <- rep(c("C", "G", "M", "T"), each = length(seq(from = 0, to = 90)))

# Combine into data frame
data_pred_zone <- data.frame(SV_length_vector_2, zone_vector_2, sex_vector_2)

#rename columns
colnames(data_pred_zone) <- c("SV_length", "zone", "sex")

# Predict using model
prediction_zone <- predict(tail_model,
                      newdata = data_pred_zone,
                      type = "response",
                      se.fit = TRUE)

# Create final data frame with prediction results
data_fig_2 <- data.frame(data_pred_zone,
                       prediction_zone$fit,
                       prediction_zone$se.fit)

# Rename columns
colnames(data_pred_zone) <- c("SV_length", "zone", "sex")

colnames(data_fig_2) <- c("SV_length", "zone", "sex", "probability", "se")

# Graph probability of a lizard tail being whole as lizard length increases, with lines colored by zone
ggplot(data_fig_2, aes(x = SV_length, y = probability)) +
  geom_line(aes(color = zone)) +
  geom_ribbon(aes(ymin = probability - se,
                  ymax = probability + se,
                  fill = zone), alpha = 0.4) +
  labs(x = "SV Length",
       y = "Probability of Having Whole Tail",
       fill = "zone") +
  theme_bw() +
  ggtitle("Probability of Whole Tail by Snout to Vent Length, by Zone")


```

## Poisson Regression

```{r}
#load data
lizard_plant <-read.csv('jrn_lizard_npp.csv')

# Research question: Do seasons and plant species cover significantly predict lizard counts?

#Analyze data (plant cover % by type, and seasons)

p4 <- ggplot(lizard_plant, aes(x = PRGL)) +
  geom_histogram(fill = "coral", color = "black") +
  labs(title = "Percent cover of PRGL plants", x = "Percent cover ", y = "Number of Observations") 

p5 <- ggplot(lizard_plant, aes(x = BOER)) +
  geom_histogram(fill = "coral", color = "black") +
  labs(title = "Percent cover of BOER plants", x = "Percent cover ", y = "Number of Observations") 

p6 <- ggplot(lizard_plant, aes(x = SCBR)) +
  geom_histogram(fill = "coral", color = "black") +
  labs(title = "Percent cover of SCBR plants", x = "Percent cover ", y = "Number of Observations") 

p7 <- ggplot(lizard_plant, aes(x = LATR)) +
  geom_histogram(fill = "coral", color = "black") +
  labs(title = "Percent cover of LATR plants", x = "Percent cover ", y = "Number of Observations")

p8 <- ggplot(lizard_plant, aes(x = season)) +
  geom_bar(fill = "magenta3", color = "black") +
  labs(title = "Lizard count by season", x = "Season ", y = "Number of Observations") 

p9 <- ggplot(lizard_plant, aes(x = BOER, y = lizard_count))+
  geom_point()

p10<- ggplot(lizard_plant, aes(x = PRGL, y = lizard_count))+
  geom_point()


p11<- ggplot(lizard_plant, aes(x = SCBR, y = lizard_count))+
  geom_point()

p12<- ggplot(lizard_plant, aes(x = LATR, y = lizard_count))+
  geom_point()

p13<- ggplot(lizard_plant, aes(x = season, y = lizard_count, color=season))+
  geom_point()

#plot
p4
p5
p6
p7
p8
p9
p10
p11
p12
p13

#LATR and SCBR seem to have stronger correlation than
```

## Analysis of Regression

```{r}
#convert necessary data in factors for GLM
#unique(lizard_plant$season)

lizard_plant$season<- factor(lizard_plant$season, levels = c("F", "S", "W"))

lizard_plant_model<- glm(lizard_count~ season+LATR+SCBR, data=lizard_plant, family= 'poisson')
                           
summary(lizard_plant_model)
plot(lizard_plant_model)
tbl_regression((lizard_plant_model))

#remove outliers and rerun GLM

lizard_plant2<-lizard_plant[-c(32,47)]
                            
lizard_plant_model2<- glm(lizard_count~ season+ LATR+SCBR, data=lizard_plant2, family= 'poisson')                  

summary(lizard_plant_model2)
plot(lizard_plant_model2)
tbl_regression((lizard_plant_model2))
                             
```

## Discussion of Regression

The results of a Poisson regression suggest that lizard counts were significantly lower in summer relative to the reference season (β = -0.80, p \< 0.001), while counts in winter did not differ significantly (β = -0.03, p = 0.62). Among the two plant species analyzed, SCBR was associated with significantly higher lizard counts (β = 0.44, p \< 0.001), whereas LATR did not have a significant effect on lizard counts (β = -0.04, p = 0.22). (Note: these coefficients are for estimation of log-transformed counts). I removed the observations on rows 32 and 47 since they appeared as outliers in the QQ plots and could skew data inaccurately (i.e it would skew data to show higher correlation of lizard count and tree species when the relationship is in reality weaker).
